---
title: "Take-Home Exercise 1"
author: "David Chiam"
description: ""
date: "1 May 2025"
date-modified: "N.A."
format: html
editor: visual
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# **1. Overview**

## **Setting the scene**

A local online media company that publishes daily content on digital platforms is planning to release an article on **demographic structures and distribution of Singapore in 2024**.

## **Our task**

In this exercise, Exploratory Data Analysis (EDA) methods and ggplot functions are used to explore:

1.  Need to write something here

2.  and here

## **2. Getting started, Data Loading and Preparation**

This section documents the process of preparing the dataset for analysis and visualization. It covers the loading of R packages, initial inspection of the dataset, and the necessary wrangling and transformation steps applied.

### **2.1 Load required packages 📦**

Before importing the data, we load several R packages from the **tidyverse** ecosystem and supporting libraries for data wrangling and visualization. These packages streamline data processing and produce publication-ready plots using `ggplot2`. This setup ensure that we have both the flexibility and power to craft expressive, publication-quality demographic visualizations for the report.

📌 **Summary of Package Roles**

| Library | Description |
|---------------------|---------------------------------------------------|
| [**tidyverse**](https://www.tidyverse.org/) | A collection of core packages designed for data science, used extensively for data preparation and wrangling |
| [**knitr**](https://yihui.org/knitr/) | Generates dynamic R markdown or Quarto reports |
| [**patchwork**](https://patchwork.data-imaginist.com/) | Combine multiple ggplots into a single layout |
| [**ggthemes**](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) | Provides extra themes (e.g. economist, wsj) |
| [**scales**](https://scales.r-lib.org/) | For data/axes label formatting and annotations for ggplot2 |
| [**ggtext**](https://wilkelab.org/ggtext/) | Supports improved and rich text rendering for ggplot2 |
| [**ggalt**](https://yonicd.github.io/ggalt/) | A compendium of ‘geoms’, ‘coords’, ‘stats’, scales and fonts for ggplots2 |
| [**ggridges**](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html) | For ridgeline plots that create the impression of a mountain range. They can be useful for visualising changes in distributions over time or space. |
| [**ggpubr**](https://github.com/kassambara/ggpubr) | For publication-ready plots. |
| [**ggextra**](https://github.com/daattali/ggExtra) | For adding marginal plots to ggplot2 |
| [**cowplot**](https://github.com/wilkelab/cowplot) | Provides precise figure alignment and plot annotation |
| [**ggnewscale**](https://eliocamp.github.io/ggnewscale/) | Enables multiple fill/colour scales in a single plot. For defining scales in ggplot2. |

We load the following R packages using the `pacman::p_load()` :

```{r}

pacman::p_load(
  tidyverse,     # Core data manipulation and ggplot2 plotting
  knitr,         # Dynamic report generation and rendering
  patchwork,     # Combines multiple ggplot2 plots
  ggthemes,      # Additional themes and visual styles
  scales,        # Formatting axes, scales, and labels
  ggtext,        # Rich text formatting in ggplot titles/labels
  ggalt,         # Dumbbell, lollipop, and other advanced geoms
  ggridges,      # Ridgeline plots to visualize distributions
  ggpubr,        # Publication-ready ggplot2 enhancements
  ggExtra,       # Marginal plots for added insight
  cowplot,       # Publication-quality figure alignment
  ggnewscale     # Allows multiple color/fill scales in one ggplot
)
```

### **2.2** Loading the Data and Taking a Glimpse of the Dataset **📂**

The dataset, *Singapore Residents by Planning Area / Subzone, Single Year of Age and Sex, June 2024*, published by the [Department of Statistics, Singapore (DOS)](https://www.singstat.gov.sg/) was loaded using the `read_csv()` function from the `readr` package in the tidyverse collection.

```{r}

# Load the dataset
df <- read_csv("Dataset/respopagesex2024.csv")

```

The dataset contains **60,424 rows** and **6 columns**. Each row represents the count of residents for a specific combination of planning area, subzone, age, sex, and year.

```{r}

glimpse(df)
```

🧾 **Variable Descriptions**

| Column | Type | Description |
|-----------|---------------------------|----------------------------------|
| `PA` | Categorical | Planning Area (e.g., Ang Mo Kio, Bedok) |
| `SZ` | Categorical | Subzone within each Planning Area |
| `Age` | Categorical (to be converted) | Age in single years (e.g., 0–100+) |
| `Sex` | Categorical | Gender: Male / Female |
| `Pop` | Continuous (Integer) | Number of residents (rounded to nearest 10) |
| `Time` | Identifier / Constant | Time period (always 2024 in this dataset) |

### **2.3** **Wrangling, Cleaning & Transformation 🧹**

To ensure the data is ready for exploratory visualization and demographic analysis, a series of wrangling and transformation steps are applied. These steps are logically grouped and reusable across multiple visualization tasks.

#### ✅ Convert **`Age`** to Integer 

The `Age` column is stored as a character string (e.g., `"100+"`), which can prevent proper sorting, numeric comparisons, and calculations. I converted it to an integer, coercing `"100+"` to `100`, and remove any missing or malformed entries.

This step ensures that age values are usable in:

-   Population pyramids
-   Median age calculations
-   Age band groupings

```{r}

df <- df %>%
  mutate(Age = as.integer(Age)) %>%  # Convert to numeric (100+ becomes NA)
  drop_na(Age)                       # Remove missing values caused by conversion


```

#### ✅ Create `AgeBand` for Grouped Analysis

Working with 101 separate age values (0–100) may clutter visualizations and dilute insight. Therefore, I bin the ages into meaningful age bands for easier demographic comparison and comparative visualization. These age bands are especially helpful for:

-   Grouped bar charts
-   Age structure comparisons across planning areas

```{r}

df <- df %>%
  mutate(AgeBand = cut(Age,
    breaks = c(0, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55,
               60, 65, 70, 75, 80, 85, 90, 95, 100, Inf),
    labels = c("0-4", "5-9", "10-14", "15-19", "20-24", "25-29",
               "30-34", "35-39", "40-44", "45-49", "50-54",
               "55-59", "60-64", "65-69", "70-74", "75-79",
               "80-84", "85-89", "90-94", "95-99", "100+"),
    right = FALSE))

```

#### ✅ Aggregate for Exploratory Plots & Key Insights

The data was group by Planning Area, Sex, and Age Band to prepare for visual comparisons across demographic segments. Using the code below, we can compute the **aggregated population totals** across:

-   **Age × Sex**: for national pyramids
-    **Planning Area**: for ranking and spatial comparisons

```{r}

# For population pyramid (Age-Sex)
pyramid_data <- df %>%
  group_by(Age, Sex) %>%
  summarise(Pop = sum(Pop), .groups = "drop")

# For top 5 planning areas
top5_pa <- df %>%
  group_by(PA) %>%
  summarise(Pop = sum(Pop), .groups = "drop") %>%
  slice_max(order_by = Pop, n = 5)

```

#### ✅ Add Region Classification Based on Planning Area

To enhance geographical analysis, we classify each Planning Area (`PA`) into one of Singapore’s five official regions (1) Central, (2) East, (3) North, (4) North-East, and (5) West. This grouping follows Singapore’s **Urban Redevelopment Authority (URA) Master Plan** and allows for cleaner regional comparisons, faceted plots, and high-level demographic summaries.

```{r}

# Updated region mapping
region_classifier <- tribble(
  ~PA, ~Region,
  # Central
  "BISHAN", "Central",
  "BUKIT MERAH", "Central",
  "BUKIT TIMAH", "Central",
  "CENTRAL AREA", "Central",
  "DOWNTOWN CORE", "Central",
  "GEYLANG", "Central",
  "KALLANG", "Central",
  "MARINA EAST", "Central",
  "MARINA SOUTH", "Central",
  "MUSEUM", "Central",
  "NEWTON", "Central",
  "NOVENA", "Central",
  "ORCHARD", "Central",
  "OUTRAM", "Central",
  "PAYA LEBAR", "Central",
  "QUEENSTOWN", "Central",
  "RIVER VALLEY", "Central",
  "ROCHOR", "Central",
  "SINGAPORE RIVER", "Central",
  "SOUTHERN ISLANDS", "Central",
  "STRAITS VIEW", "Central",
  "TANGLIN", "Central",
  "TOA PAYOH", "Central",

  # East
  "BEDOK", "East",
  "CHANGI", "East",
  "CHANGI BAY", "East",
  "PASIR RIS", "East",
  "TAMPINES", "East",

  # North
  "CENTRAL WATER CATCHMENT", "North",
  "MANDAI", "North",
  "SEMBABWANG", "North",
  "SIMPANG", "North",
  "SUNGEI KADUT", "North",
  "WOODLANDS", "North",
  "YISHUN", "North",

  # North-East
  "ANG MO KIO", "North-East",
  "HOUGANG", "North-East",
  "NORTH-EASTERN ISLANDS", "North-East",
  "PUNGGOL", "North-East",
  "SELETAR", "North-East",
  "SENGKANG", "North-East",
  "SERANGOON", "North-East",

  # West
  "BOON LAY", "West",
  "BUKIT BATOK", "West",
  "BUKIT PANJANG", "West",
  "CHOA CHU KANG", "West",
  "CLEMENTI", "West",
  "JURONG EAST", "West",
  "JURONG WEST", "West",
  "LIM CHU KANG", "West",
  "PIONEER", "West",
  "TENGAH", "West",
  "TUAS", "West",
  "WESTERN ISLANDS", "West",
  "WESTERN WATER CATCHMENT", "West"
)

# Join with main dataset
df <- df %>%
  mutate(PA = str_trim(str_to_upper(PA))) %>%
  left_join(region_classifier, by = "PA")

```

#### ✅ Compute the Weighted Median Age

As the simple median of the `Age` column would be misleading, as it treats each age as equally populated. Instead, we calculate a **weighted median** that reflects **actual population counts** per age to give a statistically more accurate median age for the **entire Singapore resident population**..

```{r}

df_median <- df %>%
  arrange(Age) %>%
  mutate(
    cum_pop = cumsum(Pop),
    total_pop = sum(Pop)
  ) %>%
  filter(cum_pop >= total_pop / 2) %>%
  slice(1)

median_age <- df_median$Age

```

This `df_summary` table forms the basis for multiple exploratory visualizations in the next section.

### 🔍 **3.1 What is the age and gender structure in selected planning areas?**

**Goal**: Understand whether the population skews young or old, and whether the gender balance is even.

**Method**: Population pyramids using mirrored horizontal bar charts, grouped by sex and age band.

```{r}
# Example structure (actual filtering by PA can be added)
ggplot(df_summary %>% filter(PA == "Ang Mo Kio"), 
       aes(x = AgeBand, y = ifelse(Sex == "Males", -Total, Total), fill = Sex)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Population Pyramid: Ang Mo Kio", y = "Population", x = "Age Band") +
  theme_minimal()

```

### 🔍 **3.2 How are age groups distributed across top 5 planning areas?**

**Goal**: Compare how each planning area’s population is composed in terms of youth, working-age, and elderly.

**Method**: Grouped bar plots faceted by planning area.

```{r}
top_5_areas <- df_summary %>%
  group_by(PA) %>%
  summarise(Pop = sum(Total)) %>%
  slice_max(order_by = Pop, n = 5) %>%
  pull(PA)

df_summary %>% 
  filter(PA %in% top_5_areas) %>%
  ggplot(aes(x = AgeBand, y = Total, fill = Sex)) +
  geom_col(position = "dodge") +
  facet_wrap(~ PA) +
  theme_minimal() +
  labs(title = "Age Band Composition Across Top 5 Planning Areas")

```

### 🔍 **3.3 Where are children (age 0–14) most concentrated?**

**Goal**: Identify family-oriented areas for policy planning (schools, childcare).

**Method**: Heatmap across Planning Area and Subzone using child population totals.

```{r}
df_children <- df %>%
  filter(Age < 15) %>%
  group_by(PA, SZ) %>%
  summarise(Children = sum(Pop), .groups = "drop")

ggplot(df_children, aes(x = reorder(SZ, Children), y = PA, fill = Children)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Child Population Density by Subzone",
       x = "Subzone", y = "Planning Area")

```

###  🔍 **3.4 What is the elderly dependency ratio across planning areas?**

**Goal**: Show where ageing populations might place pressure on the working-age base.

**Method**: Compute and plot the ratio of elderly (65+) to working-age (20–64) residents.

```{r}

df_ratio <- df %>%
  mutate(AgeGroup = case_when(
    Age >= 65 ~ "Elderly",
    Age >= 20 & Age < 65 ~ "Working",
    TRUE ~ "Other"
  )) %>%
  group_by(PA, AgeGroup) %>%
  summarise(Pop = sum(Pop), .groups = "drop") %>%
  pivot_wider(names_from = AgeGroup, values_from = Pop) %>%
  mutate(DependencyRatio = Elderly / Working)

ggplot(df_ratio, aes(x = reorder(PA, DependencyRatio), y = DependencyRatio)) +
  geom_col(fill = "darkred") +
  coord_flip() +
  labs(title = "Elderly Dependency Ratio by Planning Area",
       x = "Planning Area", y = "Elderly per Working-age Resident") +
  theme_minimal()


```

### 🔍 **3.5 Are there sex imbalances across age bands?**

**Goal**: Detect unusual gender differences by age.

**Method**: Line plots or bar charts of sex ratios by age band, faceted by region or large planning area.

```{r}
df_ratio_by_age <- df %>%
  group_by(AgeBand, Sex) %>%
  summarise(Pop = sum(Pop), .groups = "drop") %>%
  pivot_wider(names_from = Sex, values_from = Pop) %>%
  mutate(SexRatio = Males / Females)

ggplot(df_ratio_by_age, aes(x = AgeBand, y = SexRatio, group = 1)) +
  geom_line(color = "steelblue") +
  geom_point() +
  labs(title = "Sex Ratio by Age Band (Males per Female)",
       x = "Age Band", y = "Sex Ratio") +
  theme_minimal()

```

### 🔍 **3.6 Which subzones are the most densely populated?**

**Goal**: Identify where population pressure may be most intense.

**Method**: Bar chart or (if spatial data available) choropleth map of total population.

```         
r
```

CopyEdit

```{r}

df_subzone_pop <- df %>%
  group_by(SZ) %>%
  summarise(Pop = sum(Pop)) %>%
  arrange(desc(Pop)) %>%
  slice_head(n = 20)

ggplot(df_subzone_pop, aes(x = reorder(SZ, Pop), y = Pop)) +
  geom_col(fill = "darkblue") +
  coord_flip() +
  labs(title = "Top 20 Most Populated Subzones", x = "Subzone", y = "Population") +
  theme_minimal()

```

### 🔍 **3.7 What are the youngest vs oldest subzones?**

**Goal**: Compare zones with the youngest and oldest resident median ages.

**Method**: Dumbbell plot using median age per subzone.

```{r}

df_median_age <- df %>%
  group_by(SZ) %>%
  summarise(MedianAge = median(Age, na.rm = TRUE)) %>%
  arrange(MedianAge)

youngest <- df_median_age %>% slice_head(n = 10)
oldest <- df_median_age %>% slice_tail(n = 10)

combined <- bind_rows(youngest, oldest)

ggplot(combined, aes(x = reorder(SZ, MedianAge), y = MedianAge)) +
  geom_point(color = "darkgreen", size = 3) +
  coord_flip() +
  labs(title = "Top 10 Youngest and Oldest Subzones",
       x = "Subzone", y = "Median Age") +
  theme_minimal()

```

```{r}

# Load spatial packages
pacman::p_load(sf, ggplot2, tidyverse, viridis)

# Read the planning area shapefile (GeoPackage or GeoJSON recommended)
planning_areas <- st_read("Dataset/MP19_PLANNING_AREA_NO_SEA_PL.kml")  # or .shp/.geojson

# Prepare population data: total by PA
pa_pop <- df %>%
  group_by(PA) %>%
  summarise(Pop = sum(Pop))

# Ensure both PA names match exactly (some cleaning may be needed)
# Join spatial and population data
map_data <- planning_areas %>%
  left_join(pa_pop, by = c("Name" = "PA"))  # Adjust column name if different

# Plot
ggplot(map_data) +
  geom_sf(aes(fill = Pop), color = "white") +
  scale_fill_viridis_c(option = "plasma", trans = "log") +
  labs(title = "Population Distribution by Planning Area, Singapore (2024)",
       fill = "Population") +
  theme_minimal()

```

```{r}

# Load required packages
pacman::p_load(sf, ggplot2, tidyverse, viridis)

# Read + fix shapefile
planning_areas <- st_read("Dataset/MP19_PLANNING_AREA_NO_SEA_PL.kml") %>%
  mutate(Name = str_trim(toupper(Name))) %>%
  st_make_valid()

# Join and calculate density
map_data <- planning_areas %>%
  left_join(pa_pop, by = c("Name" = "PA")) %>%
  mutate(area_km2 = as.numeric(st_area(geometry)) / 10^6,
         density = Pop / area_km2)

# Plot
map_plot <- ggplot(map_data) +
  geom_sf(aes(fill = density), color = "white") +
  geom_sf_text(aes(label = Name), size = 3) +
  scale_fill_viridis_c(option = "plasma", trans = "log", na.value = "grey90") +
  labs(title = "Population Density by Planning Area (2024)",
       subtitle = "Residents per square kilometre",
       fill = "Density\n(per km²)") +
  theme_void()


```

```{r}




# Compute median age (weighted)
df_ordered <- df %>%
  arrange(Age) %>%
  mutate(cum_pop = cumsum(Pop),
         total_pop = sum(Pop)) %>%
  filter(cum_pop >= total_pop / 2) %>%
  slice(1)

median_age <- df_ordered$Age

# Population Pyramid Data
pyramid_data <- df %>%
  group_by(Age, Sex) %>%
  summarise(Pop = sum(Pop), .groups = "drop") %>%
  mutate(Pop = if_else(Sex == "Males", -Pop, Pop))

# Top 5 Planning Areas
top5_pa <- df %>%
  group_by(PA) %>%
  summarise(Pop = sum(Pop), .groups = "drop") %>%
  slice_max(order_by = Pop, n = 5)

# Plot 1: Population Pyramid
p1 <- ggplot(pyramid_data, aes(x = Pop, y = Age, fill = Sex)) +
  geom_col(width = 1) +
  geom_hline(yintercept = median_age, linetype = "dashed", color = "black", linewidth = 1) +
  scale_x_continuous(labels = abs, name = "Population") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Singapore Population Pyramid (2024)",
       subtitle = paste("Median Age:", median_age)) +
  theme_minimal() +
  theme(legend.position = "top")

# Plot 2: Top 5 Planning Areas
p2 <- ggplot(top5_pa, aes(x = reorder(PA, Pop), y = Pop)) +
  geom_col(fill = "orchid") +
  coord_flip() +
  labs(title = "Top 5 Planning Areas by Population",
       x = "Planning Area", y = "Population") +
  theme_minimal()

# Combine using patchwork
p1 + p2 + plot_layout(widths = c(2, 1))
```
