---
title: "Take-Home Exercise 1 (Part 2)"
author: "Liu Chih Yuan & David Chiam"
date: "April 29, 2025"
date-modified: "10 May 2025"
format: html
execute: 
  eval: true        # run code 
  echo: true         # show code 
  warning: false     #warning msg wont come out
  freeze: true      #speed up rendering time
---

::: callout-note
This Netlify page was originally designed and produced by [LIU CHIH-YUAN](https://cabbage-wonderland.netlify.app/takehome_ex/takehome_ex01/takehome_ex01) for his Take-Home Assignment 1.

I will be providing a critic on three good design principles and three areas for further improvement on his work. I will also preare the makeover version of the data visualisation.
:::

# Exploratory Analysis of Age and Gender Distribution Across Singapore Planning Areas (2024)

## 1. Overview

The 2024 dataset from Singapore' Department of Statistics provides resident population data by planning area, subzone, age, and sex.

### 1.1 Objective

To perform structured exploratory data analysis to uncover insights on demographic distribution across regions.

## 2. Getting Started

### 2.1 Load Packages

```{r}
pacman::p_load(tidyverse, ggrepel, ggthemes, patchwork, ggridges, scales)
```

The R packages used in this EDA are as follows:

1.  `tidyverse` core R package for data science (contains essential packages such as ggplot2)
2.  `ggrepel` for ggplot2 to repel overlapping text labels
3.  `ggthemes` extra ggplot themes
4.  `patchwork` combine ggplot
5.  `ggridges` for ridgeline plots
6.  `scales` customer number formatting

### 2.2 Import Data

```{r}
df <- read.csv("Dataset/respopagesex2024.csv")
```

```{r}
common_theme <- theme_minimal(base_size = 16) +
  theme(
    axis.text = element_text(size = 18),
    axis.title = element_text(size = 20),
    plot.title = element_text(size = 18, face = "bold"),
    legend.text = element_text(size = 17),
    legend.title = element_text(size = 16)
  )
```

#### 2.2.1 Check Missing Values

```{r}
glimpse(df)
colSums(is.na(df))
```

### 2.3 Data Processing

#### 2.3.1 Checking Column Types

```{r}
str(df)
```

We see column `Age` is "chr" (character) instead of numeric, let's check why by finding the unique value of column `Age`

```{r}
sort(unique(df$Age))
```

Here most likely it's the `90_and_over` causing it to be a `str` not `int`

#### 2.3.2 Creating a Numeric Age Column

```{r}
df <- df %>%
  mutate(
    AgeNum = suppressWarnings(
      ifelse(Age == "90_and_Over", 90, as.numeric(Age))
    )
  )
```

Since we observed people over 90 years old are categorized `90_and_above` instead of actual numbers, for the ease of plotting we hereby create a new column `AgeNum`

#### 2.3.3 Creating Age Grouping

```{r}
df <- df %>%
  mutate(
    AgeGroup = case_when(
      AgeNum <= 12 ~ "Child",
      AgeNum <= 24 ~ "Youth",
      AgeNum <= 64 ~ "Adult",
      TRUE ~ "Senior"
    )
  )
```

We create a new column `AgeGroup` for future EDA purposes

```{r}
str(df)
```

## 3. Exploratory Data Analysis

### EDA 1: Population by Age / Age Group

```{r fig.height=8, fig.width=10}
p1 <- ggplot(df, aes(x = AgeNum, y = Pop)) +
  stat_summary(fun = sum, geom = "bar", fill = "steelblue") +
  labs(title = "Total Population by Age", x = "Age", y = "Population") +
  scale_y_continuous(labels = label_comma()) +
  common_theme

p2 <- df %>%
  group_by(AgeGroup) %>%
  summarise(Pop = sum(Pop)) %>%
  ggplot(aes(x = AgeGroup, y = Pop, fill = AgeGroup)) +
  geom_bar(stat = "identity") +
  labs(title = "Population by Age Group", y = "Population") +
  scale_y_continuous(labels = label_comma()) +
  common_theme

(p1 / p2) + plot_layout(heights = c(1.2, 1))
```

**Insights:**

1.  Most residents fall between **ages 25 to 54**
2.  **Youth population is shrinking**, suggesting long-term labor sustainability issues
3.  **Senior population (65+) rising**, indicating growing need for eldercare and aging population

------------------------------------------------------------------------

::: callout-note
## 3 Good Design Principles

1.  **Good Data-Appropriate Chart Type (Total Population by Age & Population by Age Group)**

    -   A Bar Chart was used to show the visualization of the population count for each discrete age. This is apprpriate for numerical variable like age or age bands. Readers can easily assess relative sizes, which support the effective visual comparison (i.e. comparing which age has a higher count).

    -   The second plot on "Population by Age Group" has banded the ages into 4 Age Groups, namely: *Adult*, *Child*, *Senior* and *Youth*. This helps to condense the chart into age categories for quick selection if one intends to zoom in to a particularly age group for analysis.

    -   [**Assessment**]{.underline}: I would give this a rather high score on the clarity scale.

2.  **Clear Labelling and Minimal Chart Junk**

```         
-   The Axis Labels were simple and understanbleable. For both plots, the Y-axis population intervals with the use of horizontal grid lines allow for clear relative reference and comparisons. Furthermore the grid line are in lighter tone of grey which does not distract the viewer.

-   With no excessive gridlines, background noise, or distracting embellishments, Chih Yuan was able to apply the principle of minimizing “non-data ink” to maintain visual clarity.

-   [**Assessment**]{.underline}: I would give this a good score on the clarity scale.
```

3.  **Functional Simplicity**

```         
-   The plot avoids excessive color schemes, 3D effects, or elaborate design, ensuring that the focus remains solely on the dataset itself. The use of colours also made the visual easier to identify age groupings which confusion, especially avoiding "colour blind" schemes.

-   It can be observed that Chich Yuan has likely prioritized clarity over unnecessary decoration is a strong point.
```
:::

------------------------------------------------------------------------

### EDA 2: Gender Analysis

```{r fig.height=8, fig.width=10}
p3 <- ggplot(df, aes(x = Sex, y = Pop, fill = Sex)) +
  stat_summary(fun = sum, geom = "bar") +
  labs(title = "Population by Gender", x = NULL, y = "Population") +
  scale_y_continuous(labels = label_comma()) +
  common_theme

p4 <- df %>%
  group_by(Sex, AgeGroup) %>%
  summarise(Pop = sum(Pop)) %>%
  ggplot(aes(x = AgeGroup, y = Pop, fill = Sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Gender Distribution by Age Group", y = "Population") +
  scale_y_continuous(labels = label_comma()) +
  common_theme

(p3 / p4) + plot_layout(heights = c(1.2, 1))
```

**Insights:**

1.  Gender balance is nearly equal overall
2.  **Female dominates** in the senior age group, likely due to higher life expectancy

### EDA 3: Population Structure by Age / Gender

```{r fig.height=10, fig.width=10}
df_pyramid <- df %>%
  filter(AgeNum <= 90) %>%
  mutate(Pop = ifelse(Sex == "Males", -Pop, Pop))

p6 <- ggplot(df_pyramid, aes(x = AgeNum, y = Pop, fill = Sex)) +
  geom_col(width = 1) +
  coord_flip() +
  labs(title = "Population Pyramid", x = "Age", y = "Population") +
  scale_y_continuous(labels = label_comma()) +
  common_theme

p7 <- df %>%
  group_by(Sex, AgeGroup) %>%
  summarise(Pop = sum(Pop)) %>%
  ggplot(aes(x = AgeGroup, y = Pop, fill = Sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Age Group Distribution by Gender", y = "Population") +
  scale_y_continuous(labels = label_comma()) +
  common_theme

(p6 / p7) + plot_layout(heights = c(1.3, 1))
```

**Insights:**

1.  Pyramid shows **narrowing base wider top**, typical for aging societies
2.  **Adults dominate** across both genders, seniors are the second largest group

## 4. Conculsion

1.  Singapore faces a demographic shift towards aging, requiring proactive planning
2.  Uneven population spread across subzones and planning ares calls for smart urban development
3.  This EDA provides clear insights and serves as baseline for policy design, urban planning, and future modelling
